Pleiotropy of expression quantitative trait loci increases with the number of regulated genes, active cell types and tissues and bound transcription factors

### 2023-01-06 ms/ms02_intro.tex
MS

% GWAS in general
Genome-wide association studies (GWAS) have allowed the association of frequent genetic variants with common traits.
%
The number of GWAS has rapidly increased since the first GWAS around 15 years ago and has resulted in comprehensive databases such as the NHGRI-EBI GWAS Catalog \citep{2007burton.barret,2018.Parkinson.Buniello}.
%
For instance, in 2019, NHGRI-EBI GWAS Catalog contained 5687 GWAS comprising 71,673 variant-trait associations from 3,567 publications \citep{2018.Parkinson.Buniello}.

% Shortcomings of GWAS
However, the interpretation of the GWAS variants is complex because of three reasons \citep{2020.Trynka.CanoGamez}.
%
First, the causal variant is hidden in regions with high linkage disequilibrium (LD).
%
Second, GWAS do not inform about the genes, cell types, and tissues involved in the phenotypes.
%
Third, GWAS variants often affect non-coding regions, where the molecular cause is more difficult to define.

% Interest in studying GWAS
Expression QTLs (eQTLs) are variants that change the expression of a given gene in a cell type or tissue.
%
eQTLs share some of the same shortcomings as GWAS variants such as the inability to identify the causal variant in regions of high LD.
%
Moreover, the distribution of eQTLs and GWAS variants in the genome are different because of negative selection \citep{2022.Pritchard.Mostafavi}.
%
Nevertheless, eQTLs are still important data sets to interpret GWAS variants, because they provide gene regulatory information about the eQTL genes and cell types and tissues \citep{2021.Li.Mu}.
%
%In addition, colocalization analysis can help prioritize causal variants.

Many GWAS have allowed observing that many genetic variants are involved in different traits \citep{2019.Posthuma.Watanabe}.
%
In that work, the analyzed loci covered 60\% of the genome, and 90\% of these loci contained associations across different traits \citep{2019.Posthuma.Watanabe}.
%
They also found that pleiotropic genes and variants were less tissue-specific in terms of gene expression and eQTLs \citep{2019.Posthuma.Watanabe}.
%
However, the question of tissue specificity of eQTLs has been analyzed in different works with conflicting conclusions \citep{2021.Li.Mu,2018.Vijayanand.Schmiedel,2017gtex.nature}.
%
In addition, the mechanism of pleiotropy is not well understood yet.

>>>>

% GWAS in general
Genome-wide association studies (GWAS) have allowed the association of frequent genetic variants with common traits.
%
The number of GWAS has rapidly increased since the first GWAS around 15 years ago and has resulted in comprehensive databases such as the NHGRI-EBI GWAS Catalog \cite{2007burton.barret,2018.Parkinson.Buniello}.
%
For instance, in 2019, NHGRI-EBI GWAS Catalog contained 5687 GWAS comprising 71,673 variant-trait associations from 3,567 publications \cite{2018.Parkinson.Buniello}.

% Shortcomings of GWAS
However, the interpretation of the GWAS variants is complex because of three reasons \cite{2020.Trynka.CanoGamez}.
%
First, the causal variant is hidden in regions with high linkage disequilibrium (LD).
%
Second, GWAS do not inform about the genes, cell types, and tissues involved in the phenotypes.
%
Third, GWAS variants often affect non-coding regions, where the molecular cause is more difficult to define.

% Interest in studying GWAS
Expression QTLs (eQTLs) are variants that change the expression of a given gene in a cell type or tissue.
%
eQTLs share some of the same shortcomings as GWAS variants such as the inability to identify the causal variant in regions of high LD.
%
Moreover, the distribution of eQTLs and GWAS variants in the genome are different because of negative selection \cite{2022.Pritchard.Mostafavi}.
%
Nevertheless, eQTLs are still important data sets to interpret GWAS variants, because they provide gene regulatory information about the eQTL genes and cell types and tissues \cite{2021.Li.Mu}.
%
%In addition, colocalization analysis can help prioritize causal variants.

Many GWAS have allowed observing that many genetic variants are involved in different traits \cite{2019.Posthuma.Watanabe}.
%
In that work, the analyzed loci covered 60\% of the genome, and 90\% of these loci contained associations across different traits \cite{2019.Posthuma.Watanabe}.
%
They also found that pleiotropic genes and variants were less tissue-specific in terms of gene expression and eQTLs \cite{2019.Posthuma.Watanabe}.
%
However, the question of tissue specificity of eQTLs has been analyzed in different works with conflicting conclusions \cite{2021.Li.Mu,2018.Vijayanand.Schmiedel,2017gtex.nature}.
%
In addition, the mechanism of pleiotropy is not well understood yet.

### 2023-01-06 ms/ms03_results.tex
MS
ENH annotate GWAS based on OLS

To annotate traits from genome-wide associations studies (GWAS) with genes and tissues,
I developed a pipeline that systematically colocalizes variants from GWAS and expression quantitative trait loci (eQTLs) (Section Methods).
%
127 eQTL studies were downloaded from the EBI eQTL catalogue, which aims to provide uniformly processed eQTLs in many
tissues and cell types \citep{2021.Alasoo.Kerimov} (Supplementary table 1).

GWAS traits were selected from the IEU OpenGWAS database based on four criteria \citep{2018.Parkinson.Buniello}.
%
The first criterion was to exclude molecular traits such as proteome or methylome.
%	
The second criterion was to include only the European population, because most
samples from the EBI eQTL catalogue belong to the European population.
%
The third criterion was to keep only well-defined medical or physiological
conditions and exclude environmental traits such "employment status" or "self-reported" medical conditions.
%
The fourth criterion was to keep only GWAS with at least 10,000 subjects, 2,000 controls and 2,000 cases.
%
These filters resulted in 418 GWAS traits (Supplementary table 2).
%
These studies resulted in 8,580 clumped leading variants with a p-value below 5e-8 from 284 GWAS (Supplementary file).
% select count(distinct gwas_id) from tophits; 284
% select count(*) from (select DISTINCT chrom, pos, rsid, ea from tophits) as sel; 8,580
% tophits goes to OSF
% colocpleio goes to OSF

My pipeline uncovered eQTL colocalizations for 5,484 or 64$\%$ of the leading GWAS variants with 7,040 eQTL genes and 127 eQTL biosamples.
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid from tophits inner join coloc on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75) as sel; 5484
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75) as sel;  7040
% select count(*) from (select distinct eqtl_id from coloc where pp_h4_abf>=0.75) as sel;  127

@@

These results is much higher than previous colocalization analyses for autoimmune diseases by previous authors that
found 23-41\% depending on the study \citep{2016.Soranzo.Chen,2017.Cotsapas.Chun,2021.Li.Mu}.
% Discuss 25% in Chun et al 2017, Nature Genetics
% Discuss 41% (eQTL + sQTLs) in Mu et al 2021, Genome Biology
% Discuss 23% in Chen et al 2016, Cell. Genetic Drivers of Epigenetic and Transcriptional Variation in Human Immune Cells

@@

First, I analysed whether there are significant differences of some effect consequences between more and less pleiotropic eQTLs.
%
I separated eQTLs according to the count of GWAS classes and computed the EBI variant effect predictor (VEP) consequence (\citep{2016.Cunningham.McLaren}).
%
I found a significant larger number of upstream, 5'-UTR, downstream and 3'-UTR eQTLs
with 2, 3 and 4 and more classes compared to eQTLs with one class (Fig. \ref{fig:vep_consequence}).
%
I also found a significant larger number of intron, and non-coding transcript eQTLs
with 3 and 4 and more classes (Fig. \ref{fig:vep_consequence}).
%
These analyses suggest that more pleiotropic eQTLs have a stronger effect on
the coding sequence and splicing regions, which might explain partly their more pleiotropic function.

@@

Cis-regulatory modules (CRMs) are non-coding genomic regions with a higher density of bound cis-regulatory modules \citep{2021.Ballester.Hammal}.
%
I found that the odds ratio of variants annotated with CRMs vs non-annotated is significantly (Fisher's exact test) higher for variants with 2 and 3 GWAS class counts compared to class count 1 (Fig. \ref{fig:freq_tf_per_variant}b).

>>>>

To annotate traits from genome-wide associations studies (GWAS) with genes and tissues,
I developed a pipeline that systematically colocalizes variants from GWAS and expression quantitative trait loci (eQTLs) (Section Methods).
%
127 eQTL studies were downloaded from the EBI eQTL catalogue, which aims to provide uniformly processed eQTLs in many
tissues and cell types \cite{2021.Alasoo.Kerimov} (Supplementary table 1).

GWAS traits were selected from the IEU OpenGWAS database based on four criteria \cite{2018.Parkinson.Buniello}.
%
The first criterion was to exclude molecular traits such as proteome or methylome.
%	
The second criterion was to include only the European population, because most
samples from the EBI eQTL catalogue belong to the European population.
%
The third criterion was to keep only well-defined medical or physiological
conditions and exclude environmental traits such "employment status" or "self-reported" medical conditions.
%
The fourth criterion was to keep only GWAS with at least 10,000 subjects, 2,000 controls and 2,000 cases.
%
These filters resulted in 418 GWAS traits (Supplementary table 2).
%
These studies resulted in 10,621 clumped leading variants with a p-value below 5e-8 from 335 GWAS (Supplementary file).
% select count(distinct gwas_id) from tophits; 335
% select count(*) from (select DISTINCT chrom, pos, rsid, ea from tophits) as sel; 10621
% tophits goes to OSF
% colocpleio goes to OSF

My pipeline uncovered eQTL colocalizations for 5,484 or 64$\%$ of the leading GWAS variants with 7,040 eQTL genes and 127 eQTL biosamples.
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid from tophits inner join coloc
% on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75) as sel; 5484
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75) as sel;  7040
% select count(*) from (select distinct eqtl_id from coloc where pp_h4_abf>=0.75) as sel;  127

@@

These results is much higher than previous colocalization analyses for autoimmune diseases by previous authors that
found 23-41\% depending on the study \cite{2016.Soranzo.Chen,2017.Cotsapas.Chun,2021.Li.Mu}.
% Discuss 25% in Chun et al 2017, Nature Genetics
% Discuss 41% (eQTL + sQTLs) in Mu et al 2021, Genome Biology
% Discuss 23% in Chen et al 2016, Cell. Genetic Drivers of Epigenetic and Transcriptional Variation in Human Immune Cells

@@

First, I analysed whether there are significant differences of some effect consequences between more and less pleiotropic eQTLs.
%
I separated eQTLs according to the count of GWAS classes and computed the EBI variant effect predictor (VEP) consequence (\cite{2016.Cunningham.McLaren}).
%
I found a significant larger number of upstream, 5'-UTR, downstream and 3'-UTR eQTLs
with 2, 3 and 4 and more classes compared to eQTLs with one class (Fig. \ref{fig:vep_consequence}).
%
I also found a significant larger number of intron, and non-coding transcript eQTLs
with 3 and 4 and more classes (Fig. \ref{fig:vep_consequence}).
%
These analyses suggest that more pleiotropic eQTLs have a stronger effect on
the coding sequence and splicing regions, which might explain partly their more pleiotropic function.

@@

Cis-regulatory modules (CRMs) are non-coding genomic regions with a higher density of bound cis-regulatory modules \cite{2021.Ballester.Hammal}.
%
I found that the odds ratio of variants annotated with CRMs vs non-annotated is significantly (Fisher's exact test) higher for variants with 2 and 3 GWAS class counts compared to class count 1 (Fig. \ref{fig:freq_tf_per_variant}b).
